ML-based approach for accelerating global search algorithm for solving multicriteria problems\thanks{This study was supported by the Russian Science Foundation, project No 21-11-00204.}

### 2022-02-08 2022/LION/main.tex
fix figures in LaTeX

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%
\usepackage{graphicx}
\usepackage{color}
\usepackage{hyperref}

@@

\begin{document}
%
\title{ML-based approach for accelerating global search algorithm for solving multicriteria problems\thanks{Supported by organization x.}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Konstantin Barkalov\orcidID{0000-0001-5273-2471} \and
Vladimir Grishagin\orcidID{0000-0002-2884-3670} \and
Evgeny Kozinov\orcidID{0000-0001-6776-0096}}
%
\authorrunning{K. Barkalov, V. Grishagin and E. Kozinov}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Lobachevsky State University of Nizhni Novgorod, Nizhni Novgorod, Russia \email{\{konstantin.barkalov,evgeny.kozinov\}@itmm.unn.ru}, \email{vagris@unn.ru}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
The paper considers a new approach to solving multicriterial optimization problems on the base of criteria scalarization, dimensionality reduction via Peano curves and efficient global search algorithm. The novelty of the approach consists in application of machine learning methods combined with utilizing a posteriori information for acceleration of the search. Effectiveness of the proposed approach has been demonstrated by means of solving a set of multiextremal multicriterial optimization problems.

\keywords{Multicriterial Problems \and Global Optimization \and Machine Learning \and Logistic Regression.}
\end{abstract}
%
%
%
\section{Introduction}
\textcolor[rgb]{1,0,0}{Machine learning (ML) methods are powerful tools that are applied in many areas of research.}
In particular, the ML methods are successfully used for solving complicated problems of computational mathematics. One of such problem classes to which ML can be applied is the class of multicriterial optimization (MCO) models. 
In this case, ML is used, as a rule, in combination with metaheuristic algorithms \cite{Subraveti2019} that in the case of multiextremal criteria lose in efficiency to deterministic methods \cite{Kvasov2018,Sergeyev2018}.

Among the most qualitative deterministic methods for solving \textcolor[rgb]{1,0,0}{these problems} the information-statistical global search algorithm \cite{Sergeyev2013,Strongin2000} can be considered. This method proposed initially for solving scalar problems was successfully extended to MCO models \cite{GergelKozinovAPI2016,Gergel2018}.

@@

The MCO problem to be considered is formulated as follows:
\begin{equation}
f(y)=(f_1 (y),f_2 (y),\dots ,f_s (y)) \to \min, \; y\in D,
\label{eq:1}
\end{equation}
\begin{equation}
D=\left\{ y \in R^N: a_i \leq y_i \leq b_i, \; 1 \leq i \leq N, \; a,b \in R \right\}.
\label{eq:2}
\end{equation}
where $f_i (y)$, $1 \leq i \leq s$, are  criteria, $y=(y_1,y_2, \dots,y_N)$ is a vector of independent variables, $N$ is the dimension of the problem. The criteria $f_i (y)$, $1 \leq i \leq s$, are supposed to be multiextremal, to be given as a ``black-box'' functions and to satisfy the Lipschitz condition
\begin{equation}
|f_i (y')-f_i (y'')| \leq L_i \|y'-y''\|, \; y',y'' \in D, \; 1 \leq i \leq s,
\label{eq:3}
\end{equation}
where $L_i>0$, $1 \leq i \leq s$, are a priori unknown Lipschitz constants.

Any effective (Pareto-optimal) variant  in which it is impossible to reduce the values of all criteria $f_i (y)$, $1 \leq i \leq s$, at once by choosing another option   can be considered as a partial solution to the MCO problem. In practice, various scalarization techniques are often used to find effective solutions \cite{Ehrgott2005,Pardalos2017,Marler2004,GergelKozinov2020}. The present study uses the minimax convolution of partial criteria that possesses good theoretical properties
\begin{equation}
\min{f(y)} \to \min_{y\in D}{\max_{1 \leq i \leq s}{(f_i (y) \lambda_i )}} ,
\label{eq:4}
\end{equation}
where $\lambda_i \geq 0$, $1 \leq i \leq s$, $\sum_{i=1}^s{\lambda_i}=1$, are numerical indicators of the significance of each criterion.

In general case, the entire set of Pareto-optimal variants is taken as a complete solution to the MCO problem. For the numerical building of an approximation of the Pareto set a number of scalar optimization problems (\ref{eq:4}) are solved with different coefficients $\lambda_i \geq 0$, $1 \leq i \leq s$, uniformly distributed
%in the space of ones.
\textcolor[rgb]{1,0,0}{in $[0,1]$.}

@@

In the course of Pareto set assessment all trial points are partitioned into two classes. This bipartition is based on belonging to the Pareto set. For these classes, a separation hyperplane is constructed in the domain of criteria values using the logistic regression \cite{Yu2011}. Examples of separated planes are shown in Fig.~\ref{fig:1}. During the choice of a new trial point, the distances of all obtained criteria values from the separated plane are calculated and the value of $R_{PS}(i)$ is built depending on these distances.

\begin{figure}[ht]
\includegraphics[width=0.8\textwidth]{fig1.png}
\caption{Separated hyperplanes in the criteria space} \label{fig:1}
\end{figure}

Hereinafter the algorithm taking into account distances to separated hyperplane and accumulated information will be called ML\_MGSA.

@@

Computational testing was carried out on the supercomputer ``Lobachevsky'' of Nizhny Novgorod State University in the environment of the Globalizer software system \cite{globalizerSystem}.

@@

ML\_MGSA solved 25 problems (\ref{eq:4}) with different convolution coefficients $\lambda$, uniformly distributed in the field of their variation. The results of the conducted experiments are shown in Table \ref{tab:1} where ML\_MGSA with $\alpha=0$ corresponds to the version without machine learning.

@@

As it follows from the results presented in Table \ref{tab:1}, ML\_MGSA demonstrates the better quality compared to other methods.

@@

In order to estimate efficiency of machine learning procedure in ML\_MGSA, this method tested in 2 variants: with ($\alpha>0$) and without ($\alpha=0$) machine learning. Each multicriterial problem was reduced to 50 scalar problems (\ref{eq:4}) with different weight coefficients $\lambda$ distributed uniformly. Other algorithm's parameters were the same. The averaged results can be found in Table \ref{tab:2}.

@@

%1 \cite{Miettinen1999}
%2 \cite{Ehrgott2005}
%3 \cite{Pardalos2017}
%4 \cite{Strongin2000}
%5 \cite{Sergeyev2013}
%6 \cite{GergelKozinovAPI2016}
%7 \cite{Gergel2018}
%8 \cite{Marler2004}
%9 \cite{GergelKozinov2020}
%10 \cite{Yu2011}
%11 \cite{globalizerSystem}
%12 \cite{Evtushenko2014}
%13 \cite{Zilinskas2015}
%14 \cite{PISA2003}

>>>>

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%
\usepackage{graphicx}
\usepackage{color}
\usepackage{hyperref}
\usepackage{ulem}

@@

\begin{document}
%
\title{ML-based approach for accelerating global search algorithm for solving multicriteria problems\thanks{This study was supported by the Russian Science Foundation, project No 21-11-00204.}}
%
\titlerunning{ML-based approach for accelerating global search algorithm}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Konstantin Barkalov\orcidID{0000-0001-5273-2471} \and
Vladimir Grishagin\orcidID{0000-0002-2884-3670} \and
Evgeny Kozinov\orcidID{0000-0001-6776-0096}}
%
\authorrunning{K. Barkalov, V. Grishagin and E. Kozinov}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Lobachevsky State University of Nizhni Novgorod, Nizhni Novgorod, Russia \email{\{konstantin.barkalov,evgeny.kozinov\}@itmm.unn.ru}, \email{vagris@unn.ru}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
The paper considers a new approach to solving multicriterial optimization problems on the base of criteria scalarization, dimensionality reduction via Peano curves and efficient global search algorithm. The novelty of the approach consists in application of machine learning methods combined with utilizing a posteriori information for acceleration of the search. Effectiveness of the proposed approach has been demonstrated by means of solving a set of multiextremal multicriterial optimization problems.

\keywords{Multicriterial Problems \and Global Optimization \and Machine Learning \and Logistic Regression.}
\end{abstract}
%
%
%
\section{Introduction}
Machine learning (ML) methods are powerful tools that are applied in many areas of research.
In particular, the ML methods are successfully used for solving complicated problems of computational mathematics. One of such problem classes to which ML can be applied is the class of multicriterial optimization (MCO) models. 
In this case, ML is used, as a rule, in combination with metaheuristic algorithms \cite{Subraveti2019} that in the case of multiextremal criteria lose in efficiency to deterministic methods \cite{Kvasov2018,Sergeyev2018}.

Among the most qualitative deterministic methods for solving these problems the information-statistical global search algorithm \cite{Sergeyev2013,Strongin2000} can be considered. This method proposed initially for solving scalar problems was successfully extended to MCO models \cite{GergelKozinovAPI2016,Gergel2018}.

@@

The MCO problem to be considered is formulated as follows:
\begin{gather}
f(y)=(f_1 (y),f_2 (y),\dots ,f_s (y)) \to \min, \; y\in D,  \label{eq:1} \\ 
D=\left\{ y \in R^N: a_i \leq y_i \leq b_i, \; 1 \leq i \leq N, \; a,b \in R \right\}.\label{eq:2} 
\end{gather}
where $f_i (y)$, $1 \leq i \leq s$, are  criteria, $y=(y_1,y_2, \dots,y_N)$ is a vector of independent variables, $N$ is the dimension of the problem. The criteria $f_i (y)$, $1 \leq i \leq s$, are supposed to be multiextremal, to be given as a ``black-box'' functions and to satisfy the Lipschitz condition
\begin{equation}
|f_i (y')-f_i (y'')| \leq L_i \|y'-y''\|, \; y',y'' \in D, \; 1 \leq i \leq s,
\label{eq:3}
\end{equation}
where $L_i>0$, $1 \leq i \leq s$, are a priori unknown Lipschitz constants.

Any effective (Pareto-optimal) variant  in which it is impossible to reduce the values of all criteria $f_i (y)$, $1 \leq i \leq s$, at once by choosing another option   can be considered as a partial solution to the MCO problem. In practice, various scalarization techniques are often used to find effective solutions \cite{Ehrgott2005,GergelKozinov2020,Marler2004,Pardalos2017}. The present study uses the minimax convolution of partial criteria that possesses good theoretical properties
\begin{equation}
\min{f(y)} \to \min_{y\in D}{\max_{1 \leq i \leq s}{(f_i (y) \lambda_i )}} ,
\label{eq:4}
\end{equation}
where $\lambda_i \geq 0$, $1 \leq i \leq s$, $\sum_{i=1}^s{\lambda_i}=1$, are numerical indicators of the significance of each criterion.

In general case, the entire set of Pareto-optimal variants is taken as a complete solution to the MCO problem. For the numerical building of an approximation of the Pareto set a number of scalar optimization problems (\ref{eq:4}) are solved with different coefficients $\lambda_i \geq 0$, $1 \leq i \leq s$, distributed uniformly.
%in the space of ones.
%\textcolor[rgb]{1,0,0}{in $[0,1]$.}

@@

In the course of Pareto set assessment all trial points are partitioned into two classes. This bipartition is based on belonging to the Pareto set. For these classes, a separation hyperplane is constructed in the domain of criteria values using the logistic regression \cite{Yu2011}. Examples of separating planes are shown in Fig.~\ref{fig:1}. During the choice of a new trial point, the distances of all obtained criteria values from the separating plane are calculated and the value of $R_{PS}(i)$ is built depending on these distances.

\begin{figure}[ht]
\center
%\includegraphics[width=0.9\textwidth]{fig1.png}
\begin{minipage}{0.45\linewidth}
\center{\includegraphics[width=0.9\linewidth]{fig1a.png}}
\end{minipage}
\begin{minipage}{0.45\linewidth}
\center{\includegraphics[width=0.9\linewidth]{fig1b.png}}
\end{minipage}

\caption{Separating hyperplanes in the criteria space} \label{fig:1}
\end{figure}

Hereinafter the algorithm taking into account distances to separating hyperplane and accumulated information will be called ML\_MGSA.

@@

Computational testing was carried out on the``Lobachevsky'' supercomputer of Nizhny Novgorod State University in the environment of the Globalizer software system \cite{globalizerSystem}.

@@

ML\_MGSA solved 25 problems (\ref{eq:4}) with different convolution coefficients $\lambda$, distributed uniformly.
%in the field of their variation.
The results of the conducted experiments are shown in Table \ref{tab:1} where ML\_MGSA with $\alpha=0$ corresponds to the version without machine learning.
As it follows from the results presented in Table \ref{tab:1}, ML\_MGSA demonstrates the better quality compared to other methods.

@@

@@

In order to estimate efficiency of machine learning procedure in ML\_MGSA, this method tested in 2 variants: with ($\alpha>0$) and without ($\alpha=0$)
\sout{machine learning} \textcolor[rgb]{1,0,0}{use of separating planes in the criteria space}. 
Each multicriterial problem was reduced to 50 scalar problems (\ref{eq:4}) with different weight coefficients $\lambda$, distributed uniformly. Other algorithm's parameters were the same. The averaged results can be found in Table \ref{tab:2}.

@@
